Modelling BCG vaccination in the UK

### 2019-07-16 rmds/08-model-development.Rmd
updated to H comments through to model development

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the direct impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have in-direct impacts, via on-wards transmission, not captured in this estimate. Both of these limitations can be over come using a dynamic transmission model (see Chapter \@ref(introduction)), this explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes via model simulation.

@@

1. **Demographic changes** - TB dynamics develop over a long timespan, because of the potential for cases to develop active TB disease many years after infection. Over these long timespans population demographics can play an important role. An approach to include demographics is to link birth and death rates so that the modeled population is static over time. This has the advantage of making it easier to identify changes that are linked to the disease dynamics. In the model presented here birth and death processes have been incorporated based on available, age-specific, data. For years with available data this has the advantage of producing demographics which match those observed in the study population, allowing for policy relevant forecasts to be made. However, for years with limited data assumptions must be made about the likely birth and death rates.

@@

1. **Additional benefits of BCG vaccination** - there is some evidence that the BCG vaccine may reduce all-cause mortality both in the general population and specifically for TB cases (Chapter \@ref(benefical-bcg-out)). There is weaker evidence that this reduction in all-cause mortality for TB cases may be associated with a reduction in TB specific mortality. This was not included in the model presented here as the evidence was not conclusive. This means the benefits of the BCG vaccine may have been underestimated.

@@

The model is stratified into 11 age groups with 5 year age groups from 0 to 49, a single age group from 50-69, and a single age group from 70 to 89. Older adults were grouped into larger age groups as they are thought to be responsible for a small amount of TB transmission and because fine scale BCG mechanisms do not need to be modelled in these age groups. Adults aged 90+ were not modeled due to large amounts of uncertainty in the demographic data and because cases in these population represent a small fraction of total TB cases (see Chapter \@ref(data)). The number of births in a given year is incorporated as a time varying parameter. The natural mortality rate is also allowed to vary with time and is stratified by age. Immigration and emigration were not included in the demographic model as reliable age stratified data was unavailable and it is unlikely that either immigration or emigration of the UK born population is a significant driver of overall population size, or structure.

@@

The vaccination model is nested into the demographic process model and therefore vaccination is possible upon entry to each modeled age group. The target age group can be varied to represent changing BCG vaccination policy. The vaccinated population is then modeled explicitly throughout all disease compartments. The primary action of the BCG vaccine is to prevent the transition from latent to active disease, this is included for both high and low risk latent cases. Waning vaccination effectiveness has been included by stratifying vaccine effectiveness by age group. The partial protection offered by BCG vaccination against initial infection has been included as a negative modifier on the protection from latent to active disease and as a modifier on the proportion of cases that are initially infected. This allows estimates of the effectiveness of BCG vaccination at preventing active TB disease in the susceptible population to be used, as these estimates have the most robust data sources. It is assumed that latently infected individuals do not gain additional protection from re-infection from the BCG vaccine. The BCG vaccine has been modeled as being partially protective for all individuals rather than as a "take" vaccine (i.e all or nothing protection). This assumption simplifies the model and will not impact the dynamics of TB transmission, assuming that protected and unprotected BCG vaccinated individuals obey the mass action assumption (See Chapter \@ref(introduction) and [@Anderson1991; @Keeling2007]).

```{r model-flow-diag, out.width = "400px", fig.scap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described.", fig.cap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described. The TB model contains the following compartments; Susceptible ($S$), high risk latent ($H$), low risk latent ($L$), active cases with pulmonary TB ($P$), active TB cases with extra-pulmonary TB only ($E$), pulmonary cases on treatment ($T_P$), and extra-pulmonary cases on treatment ($T_E$). The vaccinated ($v$) and unvaccinated ($u$) populations are represented by $k$, such that $k = u,v$. Age stratification is represented by $a$ (where $a = 1, 2, ...,11$) in the disease model and the $0,\ 1,\ 2,\ 3$ subscripts in the demographic model. Five year age groups are modelled (i.e $0-4$, $5-9$, $10-14$, ...) up to 49 years old, with a single age group for those aged 50-69 years old and those aged 70-89 years old. Individuals aged $90+$ are not explicitly modelled. In the demographic and vaccination model the A compartment represents the demographic processes modelled in all population compartments except for the vacinated and unvaccinated susceptible populations. Solid arrows represent transition rates within the modelled populations and dashed arrows represent transition rates into, or out of the modelled populations (i.e birth and death processes).", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "model_diagrams", "tb_model_flow.png"))
```

@@

\begin{equation}
\frac{dE^{kd}_a}{dt} =  (1 - \Upsilon_a)(1 - \alpha^k_a)(\epsilon^a_HH^{k}_{a}  + \epsilon^a_LL^{k}_{a}) + \zeta_a T^{k}_a - \nu_a^E E^{k}_a - \mu^E_a E^{k}_a
  (\#eq:extra-pul-model)
\end{equation}

@@

Where the unvaccinated ($u$) and vaccinated ($v$) populations are represented by $k = u,v$ and age groups are represented by $a=0,1,2,3, ... 11$. The disease model parameters are defined as follows: $\lambda_a S^k_a$ is the force of infection; $\epsilon_a$ is the rate of activation from each latent population; $\kappa_a$ is the rate of transition into the low risk latent population; $\nu_a$ is the rate of starting treatment; $\delta$ is the protection from re-infection conferred by prior latent infection; $\Upsilon_a$ is the proportion of cases that develop pulmonary TB, with or without extra-pulmonary TB; $\mu^{P,E}_a$ is the mortality from active pulmonary ($P$) and extra-pulmonary ($E$) TB; $\zeta_a$ is the rate of treatment failure; $\phi_a$ is the rate of successful treatment; $\alpha_a$ is the age-specific effectiveness of the BCG vaccine at preventing active TB disease; and $\chi^k_a$ is the age-specific protection inferred due to vaccination from initial infection. In the unvaccinated population $\alpha = 0$ and $\chi = 0$. Parameters with an $a$ subscript, or superscript, are age-stratified.

@@

```{r tb-time-treat, fig.scap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-59) older adults (15-59). The time from symptom onset to starting treatment is shorter for cases with pulmonary TB cases across age groups, with younger cases starting treatment more rapidly than older cases.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_to_treatment.png"))
```

@@

```{r tb-treat-time-succ, fig.scap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-59) older adults (15-59). There is little evidence that the time to successful treatment differs between pulmonary and extra-pulmonary cases only but older cases appear to have a high likelihood of longer treatment times.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_cure.png"))
```

@@

The POLYMOD survey, which was conducted between May 2005 and September 2006, asked 7,290 participants across eight European countries (Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, the Netherlands, and Poland) about the number of unique contacts on a randomly assigned day of the week. Survey participants were recruited to be broadly representative of the population in terms of geographical spread, age, and sex. Children and adolescents were deliberately over-sampled due to the important role they typically play in the transmission of infectious diseases. Contacts were defined as either physical (skin-to-skin contact) or as nonphysical (two-way conversation of 3 or more words in the presence of an individual but without physical contact). The age and gender of contacts was recorded as was the duration and location of the contact event. The locations were stratified into: home; school; work; transport; leisure; and other. In total 97,904 contacts were recorded, with both physical and nonphysical contacts showing large amounts of assortativity by age.

@@

The final mean contact matrix is visualised in Figure \@ref(fig:contact-tile-plot), along with the normalised standard deviation. It is clear that the POLYMOD mixing is highly assortative with the majority of contacts occurring between those close to the same age. The highest number of contacts were between children and young adults (between 5 and 20), with the number of within age groups contacts reducing as age increased. There was some outside age group mixing for all age groups with a large amount of mixing between children and middle aged adults (i.e parents and children). There was some uncertainty for all contact rates with the minimum normalised standard deviation being 10% of mean contact rates. Contact rates between older adults and children were highly uncertain and contact rates for older adults were also generally more uncertain.

```{r contact-tile-plot, fig.scap = "a.) Mean contact rates and the b.) normalised standard deviation (\\%) of 1000 boostrapped samples of  social contacts from the POLYMOD social contact survey using 5 year age groups up 49 years old and then a single group for 50-69 year olds.", fig.cap = "a.) Mean contact rates and the b.) normalised standard deviation (\\%) of 1000 boostrapped samples of  social contacts from the POLYMOD social contact survey using 5 year age groups up 49 years old and then a single group for 50-69 year olds. Mixing is highly assortative by age with children and young adults representing the majority of contacts. There is also evidence of mixing between children and middle age adults with older children mixing with progressivly older adults. Contact rates in older adults are highly uncertain, with the most uncertainty in mixing between older adults and young children.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "contact_matrix.png"))
```

@@

The effectiveness of the BCG vaccine is usually estimated using it's effectiveness at reducing the incidence of active TB cases in a susceptible population. In the model outlined in this chapter the action of the BCG vaccine has been split into its main effect of reducing the rate of latent TB cases developing active disease and its secondary effect of reducing the likelihood of initial infection. There are few estimates of the effectiveness of the BCG vaccine at preventing active TB in cases that are already latently infected and where these estimates do exist they are not stratified by time since vaccination, or age at vaccination.[@Roy2014] As the overall effectiveness ($\alpha^T_a$) of the BCG vaccine can be estimated from the combined effectiveness at preventing initial infection ($\chi_a$) and then preventing activation in latently infected individuals ($\alpha_a$) using the following equation,

@@

There is strong evidence that the overall effectiveness of the BCG vaccine reduces over time.[@Abubakar2013; @Mangtani2017] For this reason the effectiveness of the BCG vaccination overall ($\alpha^T_a$) has been stratified by the time since vaccination (by 5 year age groups). This step-wise approach has been chosen as the majority of studies report estimates for these groups and the precise functional form of the reduction in protection is unknown. For 0-5, and 5-10, years since vaccination estimates of the effectiveness of the BCG vaccine were extracted from the MRC trial.[@Hart1972] Using the published data, Poisson regression was used to estimate Rate Ratios and 95% confidence intervals. For 10-29 years after vaccination Rate Ratio estimates from a more recent case control cohort study in the UK born vaccinated at school-age have been used.[@Mangtani2017] Table \@ref(tab:bcg-eff-est) details the estimated effectiveness for each five yearly band after initial vaccination. I have assumed that the BCG vaccine is equally effective regardless of when the vaccine is given as there is no evidence that protection reduces when given to older age groups in England.

@@

#### Demographic model parameters

@@

The number of births is incorporated into the demographic model as a time varying, noisy, parameter ($\omega(t)$). It is parameterised from the data published by the Office for National Statistics (ONS), with the available data covering all years modeled. The ONS publishes the recorded number of births in England each year starting from 1929 through to 2015, with projections avaiable through to 2101 (Figure \@ref(fig:births_england)). As there is some uncertainty as to the number of births in each year I included normally distributed noise with a standard deviation of 5% of annual births.

@@

Model simulations were begun in 1931 due to the availability of population data from the 1931 census and because data on live births was only available until 1929. The demographic model is initialised using the age grouped 1931 census data, with the assumption that the entirety of the population is UK born. Initially it is assumed that there is no BCG vaccination and no TB treatment. TB treatment is assumed to begin in 1952 with the discovery of isoniazid and BCG vaccination begins at school-age in 1953. BCG vaccination coverage is assumed to vary randomly over the time horizon of the model but to have the same distribution at all time points. The assumed distribution is normal with a mean of 80% and a standard deviation of 5%. The duration with active TB is assumed to decrease from the introduction of treatment in 1952 through to 1990 when it is assumed that detection rates were equivalent to those seen today.

@@

Dowdy et al. presented a data wish list for evidence base decision making using TB models, which maybe used to assess the usefulness of a TB model for policy makers. The data requirements included: the rate of TB transmission; probability of developing active disease after an initial infection; the rate of activation amongst cases with risk factors; protection afforded by latent TB infection; the duration of infectiousness; treatment success; and the rate of spontaneous recovery. The model presented here fulfills the majority of these criteria. The rate of TB transmission is parameterised using previously published estimates of the effective contact rate in England,[@Vynnycky1999] this parameterisation will be refined in the following chapter using incidence data from the ETS. The probability of developing active TB has been sourced from recently published modelling work that fit a model of TB transmission to contact data in low incidence countries,[@Ragonnet2017] and is stratified by age as considered important by Dowdy et al. The rate of activation amongst cases with risk factors has not been included as it has been assumed that the proportion of UK born cases in the ETS with risk factors such as HIV is low. The duration of infectiousness, and treatment success have been parameterised using the ETS, although this approach is limited by possible reporting biases in the data available. The rate of spontaneous recovery has not been modeled as it is assumed that individuals are likely to be notified before clearing TB and are also likely to rapidly be started on TB treatment. This assumptions is likely to be valid as England has a robust national health service and a strong notification framework for TB. The protection afforded by latent TB infection has been included using the most recent literature sources available. All other parameters have been parameterised using the ETS were possible and otherwise from the most robust literature sources available. In particular the effectiveness of the BCG vaccine has been parameterised using data from studies that took place in England, where available, and both the protection from initial infection and the protection from developing active disease in those latently infected has been included along with estimates of the reduction in protection over time.

The semi-stochastic transmission dynamic model of TB transmission and BCG vaccination presented in this chapter provides a detailed overview of the features required to reproduce the observed epidemiology of TB in England. The model was robustly parameterised using routine surveillance data where available and otherwise using literature sources. The assumptions required by the model can be explored by fitting the model to observed data and assessing the goodness of fit. This is the focus of the next chapter. In addition the model may also be used to explore the impact of current and historic BCG vaccination policy, both in the observed data and projected into the future. This is explored in the final chapter of this thesis.

>>>>

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have in-direct impacts, via on-wards transmission, not captured in this estimate. Both of these limitations can be over come using a dynamic transmission model (see Chapter \@ref(introduction)), this explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes via model simulation.

@@

1. **Demographic changes** - TB dynamics develop over a long timespan, because of the potential for cases to develop active TB disease many years after infection. Over these long timespans population demographics can play an important role. An approach to include demographics is to link birth and death rates so that the modeled population is static over time. This has the advantage of making it easier to identify changes that are linked to the disease dynamics. In the model presented here birth and death processes have been incorporated based on available, age-specific, data. For years with available data this has the advantage of producing demographics which match those observed in the study population, allowing for policy relevant forecasts to be made. However, for years with limited data assumptions must be made about the likely birth and death rates (see \@ref(demo-model-parameters)).

@@

1. **Additional benefits of BCG vaccination** - there is some evidence that the BCG vaccine may reduce all-cause mortality both in the general population and specifically for TB cases (Chapter \@ref(beneficial-bcg-out)). There is weaker evidence that this reduction in all-cause mortality for TB cases may be associated with a reduction in TB specific mortality. This was not included in the model presented here as the evidence was not conclusive. This means the benefits of the BCG vaccine may have been underestimated.

@@

The model is stratified into 11 age groups with 5 year age groups from 0 to 49, a single age group from 50-69, and a single age group from 70 to 89. Older adults were grouped into larger age groups as they are thought to be responsible for a small amount of TB transmission and because fine scale BCG mechanisms do not need to be modelled in these age groups. Adults aged 90+ were not modeled due to large amounts of uncertainty in the demographic data and because cases in this population represent a small fraction of total TB cases (see Chapter \@ref(data)). The number of births in a given year is incorporated as a time varying parameter. The natural mortality rate is also allowed to vary with time and is stratified by age. Immigration and emigration were not included in the demographic model as reliable age stratified data was unavailable and it is unlikely that either immigration or emigration of the UK born population is a significant driver of overall population size, or structure.

@@

The vaccination model is nested into the demographic process model and therefore vaccination is possible upon entry to each modeled age group. The target age group can be varied to represent changing BCG vaccination policy. The vaccinated population is then modeled explicitly throughout all disease compartments. The primary action of the BCG vaccine is to prevent the transition from latent to active disease, this is included for both high and low risk latent cases. Waning vaccination effectiveness has been included by stratifying vaccine effectiveness by age group. The partial protection offered by BCG vaccination against initial infection has been included as a modifier on the protection from latent to active disease and as a modifier on the proportion of cases that are initially infected. This allows estimates of the effectiveness of BCG vaccination at preventing active TB disease in the susceptible population to be used, as these estimates have the most robust data sources. It is assumed that latently infected individuals do not gain additional protection from re-infection from the BCG vaccine. The BCG vaccine has been modeled as being partially protective for all individuals rather than as a "take" vaccine (i.e all or nothing protection). This assumption simplifies the model and will not impact the dynamics of TB transmission, assuming that protected and unprotected BCG vaccinated individuals obey the mass action assumption (See Chapter \@ref(introduction) and [@Anderson1991; @Keeling2007]).

```{r model-flow-diag, out.width = "400px", fig.scap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described.", fig.cap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described. The TB model contains the following compartments; Susceptible ($S$), high risk latent ($H$), low risk latent ($L$), active cases with pulmonary TB ($P$), active TB cases with extra-pulmonary TB only ($E$), pulmonary cases on treatment ($T_P$), and extra-pulmonary cases on treatment ($T_E$). The vaccinated ($v$) and unvaccinated ($u$) populations are represented by $k$, such that $k = u,v$. Age stratification is represented by $a$ (where $a = 1, 2, ...,11$) in the disease model and the $0,\ 1,\ 2,\ 3$ subscripts in the demographic model. Five year age groups are modelled (i.e $0-4$, $5-9$, $10-14$, ...) up to 49 years old, with a single age group for those aged 50-69 years old and those aged 70-89 years old. Individuals aged $90+$ are not explicitly modelled. In the demographic and vaccination model the A compartment represents the demographic processes modelled in all population compartments except for the vaccinated and unvaccinated susceptible populations. Solid arrows represent transition rates within the modelled populations and dashed arrows represent transition rates into, or out of the modelled populations (i.e birth and death processes).", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "model_diagrams", "tb_model_flow.png"))
```

@@

\begin{equation}
\frac{dE^{kd}_a}{dt} =  (1 - \Upsilon_a)(1 - \alpha^k_a)(\epsilon^a_HH^{k}_{a}  + \epsilon^a_LL^{k}_{a}) + \zeta_a T^{k}_{Ea} - \nu_a^E E^{k}_a - \mu^E_a E^{k}_a
  (\#eq:extra-pul-model)
\end{equation}

@@

Where the unvaccinated ($u$) and vaccinated ($v$) populations are represented by $k = u,v$ and age groups are represented by $a=0,1,2,3, ... 11$. The disease model parameters are defined as follows: $\lambda_a$ is the force of infection; $\epsilon_a$ is the rate of activation from each latent population; $\kappa_a$ is the rate of transition into the low risk latent population; $\nu_a$ is the rate of starting treatment; $\delta$ is the protection from re-infection conferred by prior latent infection; $\Upsilon_a$ is the proportion of cases that develop pulmonary TB, with or without extra-pulmonary TB; $\mu^{P,E}_a$ is the mortality from active pulmonary ($P$) and extra-pulmonary ($E$) TB; $\zeta_a$ is the rate of treatment failure; $\phi_a$ is the rate of successful treatment; $\alpha_a$ is the age-specific effectiveness of the BCG vaccine at preventing active TB disease; and $\chi^k_a$ is the age-specific protection inferred due to vaccination from initial infection. In the unvaccinated population $\alpha = 0$ and $\chi = 0$. Parameters with an $a$ subscript, or superscript, are age-stratified.

@@

```{r tb-time-treat, fig.scap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-69) older adults (70-89). The time from symptom onset to starting treatment is shorter for cases with pulmonary TB cases across age groups, with younger cases starting treatment more rapidly than older cases.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_to_treatment.png"))
```

@@

```{r tb-treat-time-succ, fig.scap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012.", fig.cap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012. Age is stratified into three groups;children (0-14), adults (15-69) older adults (70-89). There is little evidence that the time to successful treatment differs between pulmonary and extra-pulmonary cases only but older cases appear to have a high likelihood of longer treatment times.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_cure.png"))
```

@@

The POLYMOD survey,[@Mossong2008] which was conducted between May 2005 and September 2006, asked 7,290 participants across eight European countries (Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, the Netherlands, and Poland) about the number of unique contacts on a randomly assigned day of the week. Survey participants were recruited to be broadly representative of the population in terms of geographical spread, age, and sex. Children and adolescents were deliberately over-sampled due to the important role they typically play in the transmission of infectious diseases. Contacts were defined as either physical (skin-to-skin contact) or as nonphysical (two-way conversation of 3 or more words in the presence of an individual but without physical contact). The age and gender of contacts was recorded as was the duration and location of the contact event. The locations were stratified into: home; school; work; transport; leisure; and other. In total 97,904 contacts were recorded, with both physical and nonphysical contacts showing large amounts of assortativity by age.

@@

The final mean contact matrix is visualised in Figure \@ref(fig:contact-tile-plot), along with the normalised standard deviation. It is clear that the POLYMOD mixing is highly assortative with the majority of contacts occurring between those close to the same age.[@Mossong2008] The highest number of contacts were between children and young adults (between 5 and 20), with the number of within age groups contacts reducing as age increased. There was some outside age group mixing for all age groups with a large amount of mixing between children and middle aged adults (i.e parents and children). There was some uncertainty for all contact rates with the minimum normalised standard deviation being 10% of mean contact rates. Contact rates between older adults and children were highly uncertain and contact rates for older adults were also generally more uncertain.

```{r contact-tile-plot, fig.scap = "a.) Mean contact rates and the b.) normalised standard deviation (\\%) of 1000 boostrapped samples of  social contacts from the POLYMOD social contact survey using 5 year age groups up 49 years old and then a single group for 50-69 year olds.", fig.cap = "a.) Mean contact rates and the b.) normalised standard deviation (\\%) of 1000 boostrapped samples of  social contacts from the POLYMOD social contact survey using 5 year age groups up 49 years old and then a single group for 50-69 year olds. Mixing is highly assortative by age with children and young adults representing the majority of contacts. There is also evidence of mixing between children and middle age adults with older children mixing with progressivly older adults. Contact rates in older adults are highly uncertain, with the most uncertainty in mixing between older adults and young children.", out.extra = "", fig.height = 8}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "contact_matrix.png"))
```

@@

The effectiveness of the BCG vaccine is usually estimated using it's effectiveness at reducing the incidence of active TB cases in a susceptible population. In the model outlined in this chapter the action of the BCG vaccine has been split into its main effect of reducing the rate of latent TB cases developing active disease and its secondary effect of reducing the likelihood of initial infection. There are few estimates of the effectiveness of the BCG vaccine at preventing active TB in cases that are already latently infected and where these estimates do exist they are not stratified by time since vaccination, or age at vaccination.[@Roy2014] The overall effectiveness ($\alpha^T_a$) of the BCG vaccine can be estimated from the combined effectiveness at preventing initial infection ($\chi_a$) and then preventing activation in latently infected individuals ($\alpha_a$) using the following equation,

@@

There is strong evidence that the overall effectiveness of the BCG vaccine reduces over time.[@Abubakar2013; @Mangtani2017] For this reason the effectiveness of the BCG vaccination overall ($\alpha^T_a$) has been stratified by the time since vaccination (by 5 year age groups). This step-wise approach has been chosen as the majority of studies report estimates for these groups and the precise functional form of the reduction in protection is unknown. For 0-4, and 5-9, years since vaccination estimates of the effectiveness of the BCG vaccine were extracted from the MRC trial.[@Hart1972] Using the published data, Poisson regression was used to estimate Rate Ratios and 95% confidence intervals. For 10-29 years after vaccination Rate Ratio estimates from a more recent case control cohort study in the UK born vaccinated at school-age have been used.[@Mangtani2017] Table \@ref(tab:bcg-eff-est) details the estimated effectiveness for each five yearly band after initial vaccination. I have assumed that the BCG vaccine is equally effective regardless of when the vaccine is given as there is no evidence that protection reduces when given to older age groups in England.

@@

#### Demographic model parameters {#demo-model-parameters}

@@

The number of births is incorporated into the demographic model as a time varying, noisy, parameter ($\omega(t)$). It is parameterised from the data published by the Office for National Statistics (ONS), with the available data covering all years modeled. The ONS publishes the recorded number of births in England each year starting from 1929 through to 2015, with projections avaiable through to 2101 (Figure \@ref(fig:births-england)). As there is some uncertainty as to the number of births in each year I included normally distributed noise with a standard deviation of 5% of annual births.

@@

Model simulations were begun in 1931 due to the availability of population data from the 1931 census and because data on live births was only available from 1929. The demographic model is initialised using the age grouped 1931 census data, with the assumption that the entirety of the population is UK born. Initially it is assumed that there is no BCG vaccination and no TB treatment. TB treatment is assumed to begin in 1952 with the discovery of isoniazid and BCG vaccination begins at school-age (15 years old) in 1953. BCG vaccination coverage is assumed to vary randomly over the time horizon of the model but to have the same distribution at all time points. The assumed distribution is normal with a mean of 80% and a standard deviation of 5%. The duration with active TB is assumed to decrease from the introduction of treatment in 1952 through to 1990 when it is assumed that detection rates were equivalent to those seen today.

@@

Dowdy et al. presented a data wish list for evidence base decision making using TB models, which may be used to assess the usefulness of a TB model for policy makers. The data requirements included: the rate of TB transmission; probability of developing active disease after an initial infection; the rate of activation amongst cases with risk factors; protection afforded by latent TB infection; the duration of infectiousness; treatment success; and the rate of spontaneous recovery. The model presented here fulfills the majority of these criteria. The rate of TB transmission is parameterised using previously published estimates of the effective contact rate in England,[@Vynnycky1999] this parameterisation will be refined in the following chapter using incidence data from the ETS. The probability of developing active TB has been sourced from recently published modelling work that fit a model of TB transmission to contact data in low incidence countries,[@Ragonnet2017] and is stratified by age as considered important by Dowdy et al. The rate of activation amongst cases with risk factors has not been included as it has been assumed that the proportion of UK born cases in the ETS with risk factors such as HIV is low. The duration of infectiousness, and treatment success have been parameterised using the ETS, although this approach is limited by possible reporting biases in the data available. The rate of spontaneous recovery has not been modeled as it is assumed that individuals are likely to be notified before clearing TB and are also likely to rapidly be started on TB treatment. This assumptions is likely to be valid as England has a robust national health service and a strong notification framework for TB. The protection afforded by latent TB infection has been included using the most recent literature sources available. All other parameters have been parameterised using the ETS where possible and otherwise from the most robust literature sources available. In particular the effectiveness of the BCG vaccine has been parameterised using data from studies that took place in England, where available, and both the protection from initial infection and the protection from developing active disease in those latently infected has been included along with estimates of the reduction in protection over time.

The semi-stochastic transmission dynamic model of TB transmission and BCG vaccination presented in this chapter provides a detailed overview of the features required to reproduce the observed epidemiology of TB in England. The model was robustly parameterised using routine surveillance data where available and otherwise using literature sources. The assumptions required by the model can be explored by fitting the model to observed data and assessing the goodness of fit. This is the focus of the next chapter. In addition the model may also be used to explore the impact of current and historic BCG vaccination policy, both in the observed data and projected into the future. This is explored in the final two chapters of this thesis.
